Visual Diagnostics of an Explainer Model -- Tools for the Assessment of LIME Explanations

### 2021-01-06 paper.Rnw
updates during meeting (final version for resubmission!)

Without an assessment of the LIME explainer model, the user is putting trust in another black-box algorithm to explain the complex black-box model. In this paper, we stress the importance of assessing the LIME explainer model. To do this, we lay out the set of claims about LIME made by \citet{ribeiro:2016}, and we provide a toolkit of visual diagnostics  for the assessment of these claims: (1) \emph{explanation scatterplots}, (2) \emph{feature heatmaps}, (3) \emph{assessment metric plots}. Using these visual diagnostics, we \hh{identify situations in which LIME does not meet its claims. In some simple examples, we  demonstrate shortfalls in LIME explanations where} the explanations are highly dependent on tuning parameter values without a clear way to identify a set of tuning parameter values that produce the "best" explanations.

@@

When quantile bins are used to simulate data, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the features in the ridge regression explainer model. The features included in the visualization are the ones selected by feature selection, the lengths of the bars represent the coefficients from the ridge regression associated with the indicator variables, while the color of a bar denotes the sign of the coefficient. This explainer model suggests that a random forest prediction of `blue' for this observation is mostly supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]} but somewhat contradicted by the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]}. %\kgc{I decided to leave the values here. I actually prefer it this way myself, but I cut out a lot of text from this subsection in other ways.}\hh{XXX }

@@

To provide a more detailed view of explanations obtained using different tuning parameter values, we consider explanation scatterplots for one prediction of interest referred to as case NM (a known non-match) shown in Figure~\ref{fig:figure-11} along with the visual explanation plots from the \emph{lime} R package. Plots for 4-equal-bins and 4-quantile-bins are included to depict tuning parameter values with good (4-equal-bins) and mediocre (4-quantile-bins) performance based on the assessment metric plot. Additional examples of explanation scatterplots including explanations for a known match and explanations produced using the kernel density simulation method are included in the supporting information.

Figure~\ref{fig:figure-11} (top left) shows that when 4-quantile-bins are applied to case NM, LIME explains that the values of mismatches, rough correlation, and distance are important and support a prediction of the class true. However, this explanation contradicts the random forest prediction which correctly assigns the observation  a non-match. The explanation scatterplot (Figure~\ref{fig:figure-11} bottom left) clarifies the reason for the highly unfaithful explanation. Case NM falls on the boundary of regions for each of these three features where the random forest predictions transition from below 0.5 to above 0.5. However, the regions used when applying LIME from the 4-quantile-bins where case NM falls contain mostly observations with random forest predictions above 0.5, leading to the unfaithful explanation.

@@

Even though an explanation will be interpretable if  the complexity of the explainer model is appropriately chosen, a lack of understanding of the explainer model could lead to an incorrect interpretation of the explanation. For example, the visualization of a LIME explanation from the \emph{lime} R package \citep{pedersen:2020} (\autoref{fig:figure-03} left; \autoref{fig:figure-11} top) \hh{provides an (over-)}simplification of the explainer model that could lead to misinterpreted LIME explanations. \kgc{Clarify?} \hh{XXX better?} Supplementing \citet{pedersen:2020}'s compact visualization of the explanation with an explanation scatterplot that shows a more detailed visualization of the explainer model promotes a more complete understanding of the explanation (\autoref{fig:figure-03} right; \autoref{fig:figure-11} bottom).

>>>>

Without an assessment of the LIME explainer model, the user is putting trust in another black-box algorithm to explain the complex black-box model. In this paper, we stress the importance of assessing the LIME explainer model. To do this, we lay out the set of claims about LIME made by \citet{ribeiro:2016}, and we provide a toolkit of visual diagnostics  for the assessment of these claims: (1) \emph{explanation scatterplots}, (2) \emph{feature heatmaps}, (3) \emph{assessment metric plots}. Using these visual diagnostics, we identify situations in which LIME does not meet its claims. In some simple examples, we  demonstrate shortfalls in LIME explanations where the explanations are highly dependent on tuning parameter values without a clear way to identify a set of tuning parameter values that produce the "best" explanations.

@@

When quantile bins are used to simulate data, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the features in the ridge regression explainer model. The features included in the visualization are the ones selected by feature selection, the lengths of the bars represent the coefficients from the ridge regression associated with the indicator variables, while the color of a bar denotes the sign of the coefficient. This explainer model suggests that a random forest prediction of `blue' for this observation is mostly supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]} but somewhat contradicted by the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]}.

@@

For a more detailed view of explanations based on different tuning parameter values, we consider explanation scatterplots for a prediction of interest, a known non-match (NM), in Figure~\ref{fig:figure-11} along with the explanation plots from the \emph{lime} R package. Plots for 4-equal-bins and 4-quantile-bins are included to depict tuning parameter values with good (4-equal-bins) and mediocre (4-quantile-bins) performance based on the assessment metric plot. Additional examples of explanation scatterplots including explanations for a known match and explanations produced using the kernel density simulation method are included in the supporting information.

Figure~\ref{fig:figure-11} (top left) shows that when 4-quantile-bins are applied to case NM, LIME finds the values of mismatches, rough correlation, and distance are important and support a prediction of the class true. However, this explanation contradicts the random forest prediction which correctly assigns the observation a non-match. The explanation scatterplot (Figure~\ref{fig:figure-11} bottom left) clarifies the reason. Case NM falls on the boundary of regions for each of these three features where the random forest predictions transition from below 0.5 to above 0.5. However, the regions used when applying LIME from the 4-quantile-bins where case NM falls contain mostly observations with random forest predictions above 0.5, leading to the unfaithful explanation.

@@

Even though an explanation will be interpretable if  the complexity of the explainer model is appropriately chosen, a lack of understanding of the explainer model could lead to an incorrect interpretation of the explanation. For example, the visualization of a LIME explanation from the \emph{lime} R package \citep{pedersen:2020} (\autoref{fig:figure-03} left; \autoref{fig:figure-11} top) provides an (over-)simplification of the explainer model that could lead to misinterpreted LIME explanations. Supplementing \citet{pedersen:2020}'s compact visualization of the explanation with an explanation scatterplot that shows a more detailed visualization of the explainer model promotes a more complete understanding of the explanation (\autoref{fig:figure-03} right; \autoref{fig:figure-11} bottom).
